---
title: |
  | \vspace{5cm} \LARGE{ISyE 6404 CP.1: Proportional Hazards Regression}
author:
- Yuan Gao, Kevin Lee, Akshay Govindaraj, 
- Yijun (Emma) Wan, Peter Williams, Ruixuan Zhang

date: "`r paste0('Date: ',Sys.Date())`"
output:
   pdf_document:
      fig_caption: true
      number_sections: false
fontfamily: mathpazo
fontsize: 11pt
header-includes:
  - \usepackage{titling}
  - \pretitle{\begin{flushleft}}
  - \posttitle{\end{flushleft}}
  - \preauthor{\begin{flushleft}}
  - \postauthor{\end{flushleft}}
  - \predate{\begin{flushleft}}
  - \postdate{\end{flushleft}}
---

\newpage
\tableofcontents

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Workload Distribution


\begin{center}
\begin{tabular}{ |l|l| } 
\hline
Team Member & Task Description  \\
\hline
Peter Williams & Code Compilation, R Debugging, Latex Formatting, Visualization  \\ 
\hline
\end{tabular}
\end{center}

\newpage

# Proportional Hazards Regression Tasks

## 1. Data Description & PH-regression
*Task: Locate a data set in the field of your interest, e.g., eCommerce, medical study, drug development, supply-chain/logistics operations, to practice the PH-regression technique. Note that we need to predict both hazard-rate and the survival function at an input x0.*
  

3) status:	event status
4) age:	in years
5) sex:	1=male, 2=female
6) disease:	disease type (0=GN, 1=AN, 2=PKD, 3=Other)
7) frail:	frailty estimate from original paper


```{r PHregression_un, echo = F, include = T, fig.align='center',fig.height=4}
#extract name of var, estimate, and p-value
#helper to extract model summaries in tabular format
get_cox_ph_results <- function(coxph_res){
  zph <- cox.zph(coxph_res)
  res_table <- data.frame(covariate = names(coxph_res$coefficients), 
            beta = as.numeric(coxph_res$coefficients),
  return(res_table)
}

variables <- c('age','sex','disease','frail')

results <- do.call('rbind',
          lapply(variables, function(x){
                  model_formula <- as.formula(paste0("Surv(time, status) ~ ",x))
          model_res <- coxph(formula = model_formula, data = kidney)
          return(get_cox_ph_results(model_res))
          }))

```









```{r PH-regression_plot, echo = F, include = T, fig.align='center',fig.height=4}
res.cox <- survfit(Surv(time, status) ~ sex, data =  kidney)
```  


res.cox <- coxph(Surv(time, status) ~ sex + frail, data =  kidney)
new_dat <- data.frame(sex = c(1), frail = c(2))
pref_df <- survfit(res.cox, newdata = new_dat)
```





# must pass indices argument so that bootstrap can randomly choose - in this case row index
# overly specified model tends to error out due to estimation error
# in this case only using age as a covariate - feel free to add others # with care

#bootstrap function - age coefficient only here
bs_fun <- function(data, indices){
	bs_dat <- data[indices,]
  	res.cox <- coxph(Surv(time, status) ~ age + sex + disease + frail , data = bs_dat)
  	return(as.numeric(res.cox$coefficients["age"]))
}

bs_res <- suppressWarnings(boot(kidney, bs_fun, R=2000))

#plot(bs_res) #if symmetric - can use percentile approach
# type options include: "norm", "basic", "perc", "stud"

bs_ci <- suppressWarnings(boot.ci(bs_res, conf = 0.90, var.t0 = NULL, type = 'perc'))

```


















Reference 3.1 introduces the Bayesian variable selection for proportional hazards regression models with right censored data where a nonparametric prior is specified for the baseline hazard rate with the use of discrete gamma process and a semi-automatic parametric informative prior specification that focuses on the observables for the regression coefficients and the model space. In addition, it proposes a Markov chain Monte Carlo method to compute the posterior model probabilities.



## 4. Implementation: Reference Procedure  

*Task: Outline steps for implementing one of the studied procedure addressed in the reference. You DO NOT need to implement them, but describe how to do it.*




As we can see from the simplified AFT model equation, AFT models (parametric as they are) are analogous to generalized linear regression models in the sense that the results of the model can be interpreted similarly to how a typical regression can be. 







## 5. Result Review  



#### PH Regression Model Fitting Code
The following code demonstrates our approach to fit each potential co-variate independently to estimate its impact in context of the survival function, utilized in task 1:  

```{r PHregression_code_appendix, echo = T, include = T}
#helper to extract model summaries in tabular format
get_cox_ph_results <- function(coxph_res){
            pvalue = 
}
variables <- c('age','sex','disease','frail')
results <- do.call('rbind',
          lapply(variables, function(x){
                  model_formula <- as.formula(paste0("Surv(time, status) ~ ",x))
          model_res <- coxph(formula = model_formula, data = kidney)
          return(get_cox_ph_results(model_res))
          }))
```

#### Bootstrap Procedure 
```{r ph_bootstrap_ca, include = T, echo = T}
# in this case only using age as a covariate below
suppressPackageStartupMessages( library(boot))
#bootstrap function - age coefficient only here
bs_fun <- function(data, indices){
	bs_dat <- data[indices,]
  	res.cox <- coxph(Surv(time, status) ~ age , data = bs_dat)
  	return(as.numeric(res.cox$coefficients["age"]))
}
bs_res <- boot(kidney, bs_fun, R=2000) # R = number of replications

# type options include: "norm", "basic", "perc", "stud"
bs_ci <- boot.ci(bs_res, conf = 0.95, var.t0 = NULL, type = 'perc')
```

*Questions?*  
*Contact: ygao390, kylee20, ywan40, agovindaraj6, pwilliams60, rzhang438 | @gatech.edu*


